---
title: "CPLN 675: Homework 3"
author: "Mimi Tran & Javier Fernandez"
date: "April 3, 2024"
  output:
  html_document:
  toc: true
toc_float: true
code_folding: hide
code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction: The Task

We have been tasked with creating a model that predicts areas vulnerable to flooding. In order to do so, we chose an area where we already had recorded flood data: Calgary, Alberta (CAN). In addition, we needed to choose a comparable city to Calgary, one that could test our model. We chose Washington DC. Both Calgary and DC have large rivers running through the cities and are large population centers. 

After choosing our cities, we came up with potential factors that might lead to flooding. Those factors are: 

- Elevation
- Impervious Surfaces
- Parks
- Low Permeable Soils (e.g. clay)
- Waterbodies

For each factor, we came up with a hypothesis: 

- Elevation: If an area has low elevation, it will most likely flood.
- Impervious Surfaces: If an area has or is made up of many impervious surfaces, it will most likely flood. 
- Parks: If an area has or is made up of many green spaces, it will most likely not flood. 
- Low Permeable Soils: If an area has or is made up of many low permeable soils, it will most likely flood.
- Waterbodies: If an area has or is made up of existing waterbodies, it will most likely flood. 

In order to test these hypotheses and build a predictive model, we conducted feature engineering, spatial lags, and a generalized linear model (GLM). The following report outlines our approach and methods, beginning with creating a training and testing set with recorded flood data in Calgary, then testing our model with predicting flooding in DC.  

Ultimately, while the DC predictive model did not perform how we would have hoped, we found promising results in our Calgary model. This draws us to conclude that Washington DC may not have been a good fit as a comparison city. For example, while Calgary is far more inland and has a drier climate, DC is relatively close to the Chesapeake Bay and tends to be more humid. A US city that might have been more comparable to Calgary might be Denver, CO. 

```{r Set Up}
# install
options(repos = "https://cloud.r-project.org/")
install.packages('caret')
install.packages('pscl')
install.packages('plotROC')
install.packages('pROC')
install.packages('sf')
install.packages('tidyverse')
install.packages('knitr')
install.packages('kableExtra')
install.packages('tigris')
install.packages('viridis')
install.packages('gstat')
install.packages('sp')
install.packages('raster')
install.packages('spatialreg')
install.packages('tidyr')
install.packages('dplyr')
install.packages(ggplot2)



# libraries
library(caret)
library(pscl)
library(plotROC)
library(pROC)
library(sf)
library(tidyverse)
library(knitr)
library(kableExtra)
library(tigris)
library(viridis)
library(gstat)
library(sp)
library(raster)
library(spdep)
library(sf)
library(tidyr)
library(dplyr)
library(ggplot2)
library(readr)


## 1. SET UP 

mapTheme <- theme(plot.title =element_text(size=12),
                  plot.subtitle = element_text(size=8),
                  plot.caption = element_text(size = 6),
                  axis.line=element_blank(),
                  axis.text.x=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks=element_blank(),
                  axis.title.x=element_blank(),
                  axis.title.y=element_blank(),
                  panel.background=element_blank(),
                  panel.border=element_blank(),
                  panel.grid.major=element_line(colour = 'transparent'),
                  panel.grid.minor=element_blank(),
                  legend.direction = "vertical", 
                  legend.position = "right",
                  plot.margin = margin(1, 1, 1, 1, 'cm'),
                  legend.key.height = unit(1, "cm"), legend.key.width = unit(0.2, "cm"))

plotTheme <- theme(
  plot.title =element_text(size=12),
  plot.subtitle = element_text(size=8),
  plot.caption = element_text(size = 6),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  # Set the entire chart region to blank
  panel.background=element_blank(),
  plot.background=element_blank(),
  #panel.border=element_rect(colour="#F0F0F0"),
  # Format the grid
  panel.grid.major=element_line(colour="#D0D0D0",size=.75),
  axis.ticks=element_blank())

```

## Uploading Data and Wrangling

We start by uploading our Calgary fishnet and features: elevation, impervious surfaces, low permeable soils, parks, and waterbodies. (Note: A fishnet is a grid made up of squares--like a net--overlaid on some geographic area. Fishnets have many uses, but our main purpose was to use the fishnet for spatial analysis and modeling.)

```{r, echo=FALSE}
# Import Calgary fishnet
cal_fish <- st_read("C:/Users/ferna/OneDrive/Documents/ArcGIS/Projects/CPLN 6750/HW3/Data_Calgary/Fishnet_Clipped/fishnet_Clip2.shp", promote_to_multi = FALSE)

# Import Calgary data features
CA_Soil <- st_read("C:/Users/ferna/OneDrive/Documents/ArcGIS/Projects/CPLN 6750/HW3/Data_Calgary/Final_Tables/CA_Soil.xlsx")
CA_Parks <- st_read("C:/Users/ferna/OneDrive/Documents/ArcGIS/Projects/CPLN 6750/HW3/Data_Calgary/Final_Tables/CA_Parks.xlsx")
CA_Impervious <- st_read("C:/Users/ferna/OneDrive/Documents/ArcGIS/Projects/CPLN 6750/HW3/Data_Calgary/Final_Tables/CA_Impervious.xlsx")
CA_Elevation <- st_read("C:/Users/ferna/OneDrive/Documents/ArcGIS/Projects/CPLN 6750/HW3/Data_Calgary/Final_Tables/CA_Elevation.xlsx")
CA_Waterbodies <- st_read("C:/Users/ferna/OneDrive/Documents/ArcGIS/Projects/CPLN 6750/HW3/Data_Calgary/Final_Tables/CA_Waterbodies.xlsx")

# Renaming and cleaning data
cal_fish <- cal_fish %>%
  rename(cell = OBJECTID_1)

CA_Soil <- CA_Soil %>%
  rename(
    cell = OBJECTID_1,
    soil_sum = SUM,
    ) %>%
  dplyr::select(-OBJECTID, -COUNT, -AREA )


CA_Parks <- CA_Parks %>%
  rename(
    cell = OBJECTID_1,
    park_sum = SUM,
    park_mean = MEAN,
    park_count = COUNT  
    ) %>%
  # Remove specific columns
  dplyr::select(-OBJECTID, -AREA, -MIN, -MAX, -RANGE, -STD, -VARIETY, -MAJORITY, -MAJORITY_COUNT, -MAJORITY_PERCENT, -MINORITY,-MINORITY_COUNT, -MINORITY_PERCENT, -MEDIAN, -PCT90 )
  

CA_Impervious <- CA_Impervious %>%
  rename(
    cell = OBJECTID_1,
    impervious_mean = MEAN,
    Impervious_majority = Impervious
    ) %>%
  dplyr::select(-OBJECTID, -COUNT, -AREA, -MAX, -MIN, -RANGE, -SUM )


CA_Waterbodies <- CA_Waterbodies %>%
  rename(
    cell = OBJECTID_1,
    Waterbodies_mean = MEAN,
  ) %>%
  dplyr::select(-OBJECTID, -AREA, -COUNT, - SUM, -MIN, -MAX, -RANGE, -STD, -VARIETY, -MAJORITY, -MAJORITY_COUNT, -MAJORITY_PERCENT, -MINORITY,-MINORITY_COUNT, -MINORITY_PERCENT, -MEDIAN, -PCT90 )

CA_Elevation <- CA_Elevation %>%
  rename(
    cell = OBJECTID_1,
    elevation_max = MAX,
    elevation_min = MIN,
    elevation_mean = MEAN,
    ) %>%
  dplyr::select(-OBJECTID, -COUNT, -AREA, -RANGE, -STD, -SUM )

# joining CSVs to Calgary fishnet by "cell" number
joined_fishcal <- cal_fish %>%
  left_join(CA_Soil, by = "cell") %>%
  left_join(CA_Parks, by = "cell") %>%
  left_join(CA_Impervious, by = "cell") %>%
  left_join(CA_Elevation, by = "cell") %>%
  left_join(CA_Waterbodies, by = "cell") 

```

After importing the Calgary fishnet and attaching our features, then we can plot the features to help visualize our data. 

```{r, echo=FALSE}
# plotting joined_fishcal, visualizing
ggplot() +
  geom_sf(data = joined_fishcal, aes(fill = Low_permeability)) +  # plot low permeability soil # logic: cells near low permeability soils will be more likely to flood
  scale_fill_gradient(low = "grey", high = "blue") +  # color
  theme_minimal()

ggplot() +
  geom_sf(data = joined_fishcal, aes(fill = Parks_Majority)) +  # plot parks # logic: cells near parks will be less likely to flood
  scale_fill_gradient(low = "grey", high = "blue") +  # color
  theme_minimal()

ggplot() +
  geom_sf(data = joined_fishcal, aes(fill = Impervious_majority)) + # plot impervious # logic: cells near impervious surfaces will be more likely to flood 
  scale_fill_gradient(low = "grey", high = "blue") +  # color
  theme_minimal()

ggplot() +
  geom_sf(data = joined_fishcal, aes(fill = elevation_min)) + # plot elevation # logic: cells near the lowest points will be more likely to flood
  scale_fill_gradient(low = "grey", high = "blue") +  # color
  theme_minimal()

ggplot() +
  geom_sf(data = joined_fishcal, aes(fill = Water_majority)) + # plot water bodies # logic: cells near water bodies will be more likely to flood
  scale_fill_gradient(low = "grey", high = "blue") +  # color
  theme_minimal()
 
```

# Conduct Spatial Lag
In order to account for any spatial dependence, we run a spatial lag on all of the features. 

``` {r}
# creating a spatial weights matrix using queen contiguity (queen =  defines neighbors as spatial units sharing a common edge)
w <- poly2nb(joined_fishcal, queen = TRUE)
w <- nb2listw(w)

# compute spatial lag with each variable
joined_fishcal$lag_soil <- lag.listw(w, joined_fishcal$Low_permeability) # low permeability soil
joined_fishcal$lag_parks <- lag.listw(w, joined_fishcal$Parks_Majority) # parks
joined_fishcal$lag_imp <- lag.listw(w, joined_fishcal$Impervious_majority) # impervious surfaces
joined_fishcal$lag_elev <- lag.listw(w, joined_fishcal$elevation_min) # elevation
joined_fishcal$lag_water <- lag.listw(w, joined_fishcal$Water_majority) # water

# 2a. PLOTTING SPATIAL LAG
# visualizing lagged variables
# Soil with Low Permeability = "Impervious Soil"
ca_soil_map <- ggplot() + 
  geom_sf(data = joined_fishcal, aes(fill = lag_soil)) +
  scale_fill_gradient(name = "Impervious Soil", low = "grey", high = "blue") +
  labs(fill = "Impervious Soil") +
  theme_minimal() # areas in blue are more likely to flood
ca_soil_map

# Presence of Parks
ca_park_map <- ggplot() + 
  geom_sf(data = joined_fishcal, aes(fill = lag_parks)) +
  scale_fill_gradient(name = "Parks", low = "grey", high = "green") +
  labs(fill = "Parks") +
  theme_minimal() # areas in green are more least likely to flood # logic: where there is a "low" count of parks, cells are more likely to flood
ca_park_map

# Presence of Impervious Surface
ca_imp_map <- ggplot() + 
  geom_sf(data = joined_fishcal, aes(fill = lag_imp)) +
  scale_fill_gradient(name = "Impervious Surfaces", low = "grey", high = "blue") +
  labs(fill = "Impervious Surfaces") +
  theme_minimal() # areas in blue are more likely to flood 
ca_imp_map

# Low Elevation
ca_elev_map <- ggplot() + 
  geom_sf(data = joined_fishcal, aes(fill = lag_elev)) +
  scale_fill_gradient(name = "Low Elevation", low = "blue", high = "grey") + 
  labs(fill = "Low Elevation") +
  theme_minimal() # areas in blue are more likely to flood # logic: areas that are low in elevation are more likely to flood
ca_elev_map # notice you can make out the river! 

# Water Bodies
ca_water_map <- ggplot() + 
  geom_sf(data = joined_fishcal, aes(fill = lag_water)) +
  scale_fill_gradient(name = "Water", low = "grey", high = "blue") + 
  labs(fill = "Water") +
  theme_minimal() # areas in blue are more likely to flood 
ca_water_map 
```

# Plotting 
```{r}
# importing inundation raster
# Specify the URL of the TIFF file
tiff_url <- "https://github.com/fdzjavier/CPLN675/raw/main/Data/rasters/Inundation_re.tif"

# Read the TIFF file into R as a raster object
inundation <- raster(tiff_url)
inundation
plot(inundation)

# upload inundation zonal statistics
inu_zonal <- st_read("C:/Users/ferna/OneDrive/Documents/ArcGIS/Projects/CPLN 6750/HW3/Data_Calgary/Final_Tables/ZonalSt_Inundation_v2.xlsx") %>%
  rename(
    cell = OBJECTID_1
  ) %>%
  dplyr::select(-OBJECTID, -AREA, -COUNT, - SUM)


# attach to fishnet
cal_inu_fish <- joined_fishcal %>%
  left_join(inu_zonal, by = "cell") 
view(cal_inu_fish)

ggplot() + 
  geom_sf(data = cal_inu_fish, aes(fill=(Flooded)), color = 'blue')

inundationPlotVariables <- 
  cal_inu_fish %>%
  as.data.frame() %>%
  dplyr::select(Flooded,Low_permeability, Parks_Majority, Impervious_majority, elevation_min, Water_majority) %>%
  gather(variable, value, -Flooded)

# Plotting Variables
test_summary <- inundationPlotVariables %>%
  group_by(Flooded, variable) %>%
  summarize(mean = mean(value))

ggplot(test_summary, aes(x = as.factor(Flooded), y = mean, fill = as.factor(Flooded))) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ variable, scales = "free_y") +
  scale_fill_manual(values = c("darkblue", "darkgreen"),
                    labels = c("Not Flooded", "Flooded"),
                    name = "") +
  labs(x = "Flooded", y = "Value")

ca_inundation <- cal_inu_fish %>%
  dplyr::select(cell, Flooded, Low_permeability, Parks_Majority, Impervious_majority, elevation_min, Water_majority)
```

# Model Building

```{r}
set.seed(3456)
trainIndex <- createDataPartition(ca_inundation$Flooded, p = .70,
                                  list = FALSE,
                                  times = 1)

floodTrain <- ca_inundation[ trainIndex,]
floodTest  <- ca_inundation[-trainIndex,]

floodModel <- glm(Flooded ~ ., 
                     family="binomial"(link="logit"), data = floodTrain %>%
                       as.data.frame() %>%
                       dplyr::select(-geometry, -cell))
summary(floodModel)
  

# Model validation for the testing dataset
classProbs <- predict(floodModel, floodTest, type="response")

hist(classProbs)
  

testProbs <- data.frame(obs = as.numeric(floodTest$Flooded),
                        pred = classProbs)

#Probability curve for the testing dataset
ggplot(testProbs, aes(x = pred, fill=as.factor(obs))) + 
  geom_density() +
  facet_grid(obs ~ .) + 
  xlab("Probability") +
  ylab("Frequency")+
  geom_vline(xintercept = .25) +
  scale_fill_manual(values = c("dark blue", "dark green"),
                    labels = c("Not Flooded","Flooded"),
                    name = "")+
  plotTheme
```


